Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs

### 2014-10-18 papers/termination/termination-esop/paper.tex
Typos

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
non-linearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.  Furthermore, our approach allows
us to \emph{simultaneously} search for proofs of \emph{non-termination},
which take the form of recurrence sets.

@@

We present a technique that treats uniformly linear and non-linear programs and
it is not restricted to finding linear ranking functions, but can
also compute lexicographic nonlinear ones. 
Our approach is constraint-based and relies on second-order formulations 
of termination and non-termination.
The obvious issue is that,
due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
To make solving our constraints tractable, we formulate termination and non-termination inside a fragment of second-order logic with restricted quantification, 
for which we have built a solver in \cite{kalashnikov}.
Our method is sound and
complete for bit-vector programs -- for any program, we find
a proof of either its termination or non-termination.

@@

\begin{definition}[Second-Order SAT]
\label{def:2sat}
 \[
  \exists S_1 \ldots S_m . Q_1 x_1 \ldots Q_n x_n . \sigma
 \]
 Where the $S_i$'s range over predicates,
the $Q_i$'s are either $\exists$ or $\forall$,
the $x_i$'s range over boolean values,
 and $\sigma$ is a quantifier-free propositional formula
 whose free variables are the $x_i$'s.  
Each $S_i$ has an associated arity $\mathrm{ar}(S_i)$
 and $S_i \in \mathbb{B}^{\mathrm{ar}(S_i)}$.  Note that 
$Q_1 x_1 \ldots Q_n x_n . \sigma$
 is an instance of first-order propositional SAT, i.e. QBF.
\end{definition}

@@

Since $L$ either terminates or does not terminate, this formula is a tautology in second-order SAT.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other, which will be exemplified by Algorithm~\ref{alg:termination} once
we have shown how to solve the formula {\bf [GT]}.

@@

\subsection{Solving the Second-Order SAT Formula}
In order to solve the second-order generalised formula {\bf [GT]}, we use the solver described in \cite{kalashnikov}.
%
%% \begin{theorem}
%%  \label{thm:synth-semicomplete}
%%  If the existential first-order theory used to express the specification $\sigma$ is decidable
%%  and the domain of inputs $X$ is finite,
%%  the second-order SAT decision procedure is semi-complete -- if a solution $(P, x_0) \models \sigma$
%%  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
%%  satisfies the specification, the algorithm may not terminate.
%% \end{theorem}
%
%
This second-order SAT solver is a semi-decision procedure -- for any
satisfiable formula, the solver is guaranteed to find a satisfying assignment to
all the second-order variables.

@@

\section{Soundness, Completeness and Complexity}
In this section, we show that $\mathcal{L}$ is expressive enough
to compute (non-)termination proofs for every bit-vector program.
By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
we then show that our analysis terminates with a valid proof for every input program.

@@

It should be noted that the comparison here is imperfect, since {\sc ARMC}
is solving a slightly different problem -- it checks whether the program
under analysis would terminate if run with unbounded integer variables,
while we are checking whether the program terminates with bit-vector
variables.  This means that {\sc ARMC}'s verdict differs from ours
in 3 cases (due to the differences between integer and bit-vector
semantics).  There are a further 7 cases where our tool is able to find a
proof and {\sc ARMC} cannot, which we believe is due to our more expressive
proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to
find a termination proof.  Of these, 2 cases have nested loops and the
third has an infinite number of terminating lassos.  This is not a problem for us,
but can be difficult for provers that enumerate lassos.

>>>>

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
nonlinearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.  Furthermore, our approach allows
us to \emph{simultaneously} search for proofs of \emph{non-termination},
which take the form of recurrence sets.

@@

We present a technique that treats linear and nonlinear programs uniformly and
it is not restricted to finding linear ranking functions, but can
also compute lexicographic nonlinear ones. 
Our approach is constraint-based and relies on second-order formulations 
of termination and non-termination.
The obvious issue is that,
due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
To make solving our constraints tractable, we formulate termination and non-termination inside a fragment of second-order logic with restricted quantification, 
for which we have built a solver in \cite{kalashnikov}.
Our method is sound and
complete for bit-vector programs -- for any program, we find
a proof of either its termination or non-termination.

@@

\begin{definition}[Second-Order SAT]
\label{def:2sat}
 \[
  \exists S_1 \ldots S_m . Q_1 x_1 \ldots Q_n x_n . \sigma
 \]
 Where the $S_i$'s range over predicates,
the $Q_i$'s are either $\exists$ or $\forall$,
the $x_i$'s range over boolean values,
 and $\sigma$ is a quantifier-free propositional formula
 whose free variables are the $x_i$'s.  
Each $S_i$ has an associated arity $\mathrm{ar}(S_i)$
 and $S_i \subseteq \mathbb{B}^{\mathrm{ar}(S_i)}$.  Note that 
$Q_1 x_1 \ldots Q_n x_n . \sigma$
 is an instance of first-order propositional SAT, i.e. QBF.
\end{definition}

@@

Since $L$ either terminates or does not terminate, this formula is a tautology in second-order SAT.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other.

@@

\subsection{Solving the Second-Order SAT Formula}
In order to solve the second-order generalised formula {\bf [GT]}, we use the solver described in \cite{kalashnikov}.
%
%% \begin{theorem}
%%  \label{thm:synth-semicomplete}
%%  If the existential first-order theory used to express the specification $\sigma$ is decidable
%%  and the domain of inputs $X$ is finite,
%%  the second-order SAT decision procedure is semi-complete -- if a solution $(P, x_0) \models \sigma$
%%  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
%%  satisfies the specification, the algorithm may not terminate.
%% \end{theorem}
%
%
For any satisfiable formula, the solver is guaranteed to find a satisfying assignment to
all the second-order variables.

@@

\section{Soundness, Completeness and Complexity}
In this section, we show that $\mathcal{L}$ is expressive enough
to capture  (non-)termination proofs for every bit-vector program.
By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
we then show that our analysis terminates with a valid proof for every input program.

@@

It should be noted that the comparison here is imperfect, since {\sc ARMC}
is solving a different problem -- it checks whether the program
under analysis would terminate if run with unbounded integer variables,
while we are checking whether the program terminates with bit-vector
variables.  This means that {\sc ARMC}'s verdict differs from ours
in 3 cases (due to the differences between integer and bit-vector
semantics).  There are a further 7 cases where our tool is able to find a
proof and {\sc ARMC} cannot, which we believe is due to our more expressive
proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to
find a termination proof.  Of these, 2 cases have nested loops and the
third has an infinite number of terminating lassos.  This is not a problem for us,
but can be difficult for provers that enumerate lassos.
